Parallel Sorting of Roughly-Sorted Sequences\\CSCI 5172 Fall '16 Project

### 2016-12-08 tex/paper.tex
updated refs and paper content, still need more results discussion

The parallelism provided by Nvidia and CUDA in this scheme can be realized by developers who develop their payloads in specific ways.  Fundamentally, programs running on an Nvidia card via CUDA operate differently than their  CPU counterparts.  In general, a line of code executing on an Nvidia card is not processed in isolation.  Rather, the Nvidia CUDA scheme provides for single instruction, multiple thread (SIMT) execution, as opposed to the usual single instruction, single data scheme (SISD) provided by general purpose CPUs.  In contrast to a single instruction accessing or updating one memory location on a single core of a general purpose CPU, a single instruction executing on an Nvidia card is executing inside multiple threads simultaneously, each of which operates on a different location in memory.

When a developer wants to take advantage of this parallelism, they structure their code such that the multiple results of the same instruction executing is a useful component of the overall program's payload.  Nvidia CUDA provides simple mechanisms by which to achieve this.

The core component of an Nvidia payload is a thread; an instance of the series of instructions in the overall CUDA payload.  Multiple threads run in a group called a block.  Blocks contain a developer-specified number of threads each.  Between threads and blocks is the concept of a warp.  Warps are collections of 32 threads executing simultaneously on the same processor core on the Nvidia card.  Usually, all threads in a warp are executing the same instruction simultaneously.  Blocks are divvyed up into multiple warps for scheduling an execution on an Nvidia GPU.  Collections of these blocks are referred to as a grid.  Finally, encompassing all of these components is the kernel.  Kernels are the series of instructions which threads, warps, blocks, and grids comprise an instance of and represent the sequence of code to actually be represented on GPU.

Nvidia provides a simple mechanism by which developers can access the parallelism provided by CUDA.  Each of the above divisions of the CUDA payload are provided with an identifier.  A tuple composed of the thread, block, and grid identifier\footnote{Concerns with warps are abstracted away from the developer.  Consequently warp identifiers are not needed and are not available.} uniquely identifies that component of the workload.  Using this information, individual threads all executing the same instruction (such as a memory from an array) can perform that instruction on distinct data (access different memory locations in the array).\cite{nvidiaguide}

Using these techniques provided a platform to develop the GPU parallel versions of sequence radius finding and subsequent roughsorting.  CUDA implementations of the RL, LR, and DM sequences were developed.  This was done by having assigning each thread a unique identifier and then, analagous to Altman and Chlebus\textquoteright work  \cite{altman89}, having teach thread manage the corresponding slot in the global unsorted array.  Each thread performed the parallel min- then max-prefix operations until the respective lists were efficiently determined ordered.  Sorting was determined by having each thread determine if its assigned slot in the LR or RL list was greater than the next element in the list, and having each thread globally assert any true (unsorted) result to global memory, available to all other threads.

Once the RL and LR lists were completed, the DM list was computed by using Altman et. al.\textquoteright s serial radius techniques in conjunction with the corresponding log \cite{altman89}.

Given many GPU processor threads may update multiple memory locations simultaneously, and likewise many memory location may be read in parallel on GPU, the underlying NVIDIA hardware that CUDA provides access to is reminiscent of CRCW PRAM.  However, GPUs differ from true CRCW PRAM in fundamental ways.  First, GPU memory is finite, while PRAM is of arbitrary size.  Furthermore, for GPU memory access times to provide the concurrent read and write characteristics reminiscent of PRAM, these memory operations must be performed on contiguous portions of GPU memory, whereas CRCW PRAM has no such ordering limitation.  Foremost in the differences between CRCW PRAM and GPU architectures however is the warp divergence concept addressed above.  What this implies for memory operations is that workloads exist which may prevent all but one thread in a warp from performing a memory operation at a particular time.  Consequently, there are practical limitations to what a GPU may achieve, and these limitations restrict the workloads which are practical to run on a GPU.  Not only must a GPU workload access constrain its memory usage within practical limitations, the accesses to that memory must be orderly, and the underlying algorithm performing those accesses must be amenable to parallelization on GPU if the advantages of GPU processing over CPU processing are to be realized\cite{dehne2010exploring}.

>>>>

The parallelism provided by Nvidia and CUDA in this scheme can be realized by developers who develop their payloads in specific ways.  Fundamentally, programs running on an Nvidia card via CUDA operate differently than their  CPU counterparts.  In general, a line of code executing on an Nvidia card is not processed in isolation.  Rather, the Nvidia CUDA scheme provides for single instruction, multiple thread (SIMT) execution, as opposed to the usual single instruction, single data scheme (SISD) provided by general purpose CPUs.  In contrast to a single instruction accessing or updating one memory location as the result of one instruction executing on a single core of a general purpose CPU, a single instruction executing on an Nvidia card is executing inside multiple threads on multiple respective cores simultaneously, each of which operates on a different location in memory.

When a developer wants to take advantage of this parallelism, they structure their code such that the multiple results of the same instruction executing with values from different memory locations is a useful component of the overall program's payload.  Nvidia CUDA provides simple mechanisms by which to achieve this.

The core component of an Nvidia payload is a thread; an instance of the series of instructions in the overall CUDA payload.  Multiple threads run in a group called a block.  Blocks contain a developer-specified number of threads each.  Between threads and blocks is the concept of a warp.  Warps are collections of 32 threads executing simultaneously on the same processor core on the Nvidia card.  Usually, all threads in a warp are executing the same instruction simultaneously. However, in situations where a conditional is evaluated by members of a warp, all threads for whom the conditional yields true will execute, and subsequently all threads with a false conditional result will execute, in a phenomenon called warp divergence.  This yields from the SIMT architecture, where all cores on which a warp executes must be processing the same instruction or temporarily switched off and not processing at all.  Blocks are divvyed up into multiple warps for scheduling and execution on an Nvidia GPU.  Collections of these blocks are referred to as a grid.  Finally, encompassing all of these components is the kernel.  Kernels are the series of instructions which threads, warps, blocks, and grids comprise an instance of and represent the sequence of code to actually be transferred to and executed on GPU.

Nvidia provides a simple mechanism by which developers can access the parallelism provided by CUDA.  Each of the above divisions of the CUDA payload are provided with an identifier.  A tuple composed of the thread, block, and grid identifiers\footnote{Concerns with warps are abstracted away from the developer.  Consequently warp identifiers are not needed and are not available.} uniquely identifies that component of the workload.  Using this information, individual threads all executing the same instruction (such as a memory read from an array) can perform that instruction on distinct data (access different memory locations in the array).\cite{nvidiaguide}

Using these techniques provided a platform to develop the GPU parallel versions of sequence radius finding and subsequent roughsorting.  In support of the above discussed parallel roughsort implementation, CUDA implementations of the RL, LR, and DM sequences were developed.  This was done by assigning each thread a unique identifier and then, analagous to Altman and Chlebus\textquoteright work, having each thread manage the corresponding slot in the global unsorted array \cite{altman89}.  Each thread performed the parallel min- then max-prefix operations until the respective lists were efficiently determined ordered.  Sorting was determined by having each thread determine if its assigned slot in the LR or RL list was greater than the next element in the list, and having each thread globally assert any true (unsorted) result to global memory, available to all other threads.

Once the RL and LR lists were completed, the DM list was computed by using Altman et. al.\textquoteright s serial radius techniques, while using the log(n) bounded techniques introduced by their RL and LR sequence determination algorithms.  However, a notable divergence from the serial version of the serial DM list algorithm was taken.  The inner loop of the serial DM list algorithm added a criteria which assured that B[i] (the ith element of the LR list) was always greater than or equal to C[i] (the ith element of the RL list).  As discussed, only the power of 2 greater than the actual value of k is actually needed for parallel roughsort to function, so this criteria was removed in the interest of efficiency and simplicity\cite{altman89}.

Finally, to determine the radius from the DM list, the maximum element needed to be found.  For this purpose, the Thrust CUDA library was used, which provides CUDA-optimized implementations of a number of C++ Standard Template Library (STL) functions\cite{thrust}.  One such STL function is the std::max\textunderscore element function.  Using the Thrust implementation of this function, maximum elements of large arrays can be found expediently on Nvidia hardware using CUDA\cite{extrema}.  Using this function yielded the maximum element of the DM list, which is k.  Further, to ensure all values quiesced during LR, RL, and DM list computations, a separate kernel was launched to perform the final DM computations.

Running these kernels yielded disappointing results, however.  For random lists composed together in such a manner as to yield a specific k value for the list\footnote{Lists which had a specific k value were generated ahead of time and then evaluated using the raidus finders to prove both correctness and provide for validity in this experimentation}, sequential CPU computations exceeded performance on GPU for all values of k beyond 6.\footnote{These results reversed when performed in a Windows environment, compiled with Visual Studio 2015 Community and run on commodity hardware.  The authors suspect this to be a false result and suspect that optimization differences on the C++ STL nth\textunderscore element function are the cause.  This begets further investigation, however}  Various performance results are reported below.

[graphs here]

The source papers theorized the converse result, so this result must then be explained.  There are several possible reasons for this which should be considered, as there are similartities but also differences between GPU and the assumed CRCW PRAM.  Given many GPU processor threads may update multiple memory locations simultaneously, and likewise many memory location may be read in parallel on GPU, the underlying NVIDIA hardware that CUDA provides access to is reminiscent of CRCW PRAM.  However, GPUs differ from true CRCW PRAM in fundamental ways.  First, GPU memory is finite, while PRAM is of arbitrary size.  Furthermore, for GPU memory access times to provide the concurrent read and write characteristics reminiscent of PRAM, these memory operations must be performed on contiguous portions of GPU memory, whereas CRCW PRAM has no such ordering limitation.  Foremost in the differences between CRCW PRAM and GPU architectures however is the warp divergence concept addressed above.  What this implies for memory operations is that workloads exist which may prevent all but one thread in a warp from performing a memory operation at a particular time.  Consequently, there are practical limitations to what a GPU may achieve, and these limitations restrict the workloads which are practical to run on a GPU.  Not only must a GPU workload access constrain its memory usage within practical limitations, the accesses to that memory must be orderly, and the underlying algorithm performing those accesses must be amenable to parallelization on GPU if the advantages of GPU processing over CPU processing are to be realized\cite{dehne2010exploring}.
